---
title: "Debating Deference: The Politicization of the Chevron Doctrine"
author: "Sam Church"
description: "Final Project for Gov 50 --- Fall 2023"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction

Chevron Deference, for a majority of Americans, is not a common topic of discussion. The result of evolving standards of statutory interpretation at the highest court in the land, it is understandable that Chevron Deference, a legal standard that results in increased flexibility regarding the authority of executive agencies, is not usually mentioned outside of the courtroom.

Recently, however, it has become increasingly utilized by democrats as a way to circumvent legislative gridlock. Because Chevron Deference, in theory, should result in the court delegating agencies broader authority to make rules, the executive branch sees it as a way to achieve broad policy objectives such as climate change regulations, reform of the financial system, and even changing standards for higher education.

Thus, it would be interesting to further analyze whether or not the court utilized Chevron deference more or less when it was largely populated by democrat-leaning or conservative-leaning justices. This is because, if democrats now are using it to circumvent a partisan legislature, it could be interesting to analyze whether or not other parties used the same tactic in the past. Understanding the influence political ideology plays on how much the Court chooses to utilize the Chevron doctrine is important now more than ever as conservatives are attempting to limit the doctrine. If conservatives typically utilize the doctrine in addition to liberals, then it will be clear that attacks on Chevron now are more ideologically motivated rather rooted in legal truth.

Therefore, a question can be proposed: does the ideological slant of the Supreme Court influence the win rate of agencies in cases citing the Chevron Deference doctrine? 

I hypothesize that agency win rate will remain relatively consistent regardless of the ideological slant of the Supreme Court. This is because the doctrine of deference is supposedly apolitical. The decision that created the doctrine was established based on precedence, and it should therefore be consistently applied regardless of the overall ideological slant of the court.

In other words, when examining data about whether or not the Supreme Court sides with an executive agency in a decision, I expect the agency win rate to remain relatively consistent. The explanatory variable of interest my project will focus on is the median justice by ideological slant. The outcome variable will be the win rate of agencies in cases that cite Chevron Deference throughout different terms. Thus, the database I will be utilizing is a list of Supreme Court cases from 1984 to 2019 where an agency attempts to interpret a law (statutory interpretation). In combination with this, I will be using a dataset that measures the ideological slant of the median Supreme Court justice for the same time period mentioned above. 

If there is a relationship between partisan slant of the Supreme Court and agency win rate in Chevron Cases my hypothesis will have been found to NOT have statistical support. If there is no relationship found, my hypothesis will have been statistically supported.

## Data

There are two interesting variables that will be analyzed: the median ideological slant of the Supreme Court for a given term and the percent of cases citing Chevon Deference an agency won in a given term. The explanatory (independent) variable will be the median ideological slant of the Supreme Court. This is a good measure of the overall ideological slant of the Supreme Court, because it indicates that there are 4 justices more conservative than them, and 4 justices more liberal than them, isolating an overall value for the ideological slant of the court. This data is the result of a measurement system known as "Martin-Quinn Scores." It uses empirically successful methods to compare decisions a justice makes and assign them an overall ideological slant, with lower numbers representing a more liberal justice. This data was collected through a survey of decisions certain justices made, and is described more in-depth in the paper where the scores were first introduced, here: https://mqscores.lsa.umich.edu/media/pa02.pdf

The dependent variable is the percent of cases citing Chevron that are ruled in favor of the agency. This can be measured by taking every case that cites Chevron, determining if the decision ruled in favor of the agency, and dividing that by the total number of cases citing Chevron in a given term. The data for this variable is a survey of cases relating to Chevron deferences, located here: https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/ZNCKL5. The variable called favors_agency can be considered in determining win rate. It is coded as a 0 if the decision favors agency, 1 if it does not, and 2 if it was mixed. To minimize confounding factors, I will only consider a 0 as a win for an agency. In addition, to avoid confounding variables regarding other aspects of cases, I will only investigate cases where the court determined Chevron Deference (one of the deference doctrines) should apply. Thus, the only cases I will evaluate are ones where Chevron Step 0 (does the Chevron framework apply) is coded as a 1 to indicate yes.


This is a cross-sectional research project. The partisan slant of the Supreme Court and the agency win rate in cases citing Chevron Deference will be analyzed at specific points in time. Specifically, they'll be analyzed during specific Supreme Court terms.


```{r}
library(ggplot2)
library(tidyverse)

deference_data <- read_csv("data/ProjectDeferenceData.csv")
court_data <- read_csv("data/court.csv")

deference_data <- deference_data |>
  mutate(
    case_winner = if_else(Decision == 0, "gov_agency", "not_agency"),
    term = Term
  ) |>
  select(term, Term, Name, Decision, Agency, ChevronCited, case_winner)


court_data <- court_data |>
  filter(term >= 1983 & term <= 2019) |>
  mutate(
    term = as.double(term)
  ) |>
  select(term, med)

final_data <- deference_data |>
  left_join(court_data)

chevron_data <- final_data |>
  group_by(term) |>
  summarize(
    median_slant = mean(med),
    agency_prop_win = mean(case_winner == "gov_agency")
  )

chevron_data <- chevron_data |>
  drop_na()

chevron_data |>
  knitr::kable(col.names = c("Term", "Ideological Slant of Median Justice", "Proportion of Agency Wins in Chevron Cases"))

```

```{r}
chevron_data |>
ggplot(mapping = aes(x = term)) +
  geom_line(aes(y = agency_prop_win, color = "Win Rate of Agencies in Chevron Cases"), se = FALSE) +
  geom_line(aes(y = median_slant, color = "Ideological Slant of SCOTUS"), se = FALSE) +
  scale_colour_manual(name ="Legend", values = c("Win Rate of Agencies in Chevron Cases" = "steelblue", "Ideological Slant of SCOTUS" = "indianred")) +
  scale_x_continuous(breaks=seq(1983,2019,by=5)) +
labs(
  title = "Ideological Slant of SCOTUS and Agency Win Rate in Cases Citing Chevron Over Time",
  x = "Year",
  y = "",
  subtitle = "Lower Position on Y-Axis Indicates a Lower Agency Win Rate and More Liberal Court",
)
  
```

## Results

```{r}
ideology_graph <- chevron_data |>
  ggplot(mapping = aes(x = median_slant, y = agency_prop_win)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Ideological Slant of Median Supreme Court Justice",
    y = "Proportion of Cases Citing Chevron an Agency Won ",
    title = "Ideological Slant of Supreme Court vs. Agency Win Rate in Chevron Cases",
    caption = "Negative Values Show Liberal Lean, Positive Values Show Conservative Lean"
  )

ideology_graph
```


```{r}
party_analysis <- lm(agency_prop_win ~ median_slant, data = chevron_data)

party_analysis
```


```{r}
library(broom)

glance(party_analysis) |>
  select(r.squared, p.value) |>
  knitr::kable(col.names = c("R Squared", "P Value"))
```





After plotting a graph of the median ideological slant of the Supreme Court on the x-axis and the percentage of cases citing Chevron that an executive agency won for each term, there is a slight positive relationship between ideological slant and percentage of cases citing Chevron that an executive agency won for each term. This is revealed by the slope of 0.09178, which demonstrates that as the ideological slant increases by 1 unit (more conservative), the agency win rate in cases citing Chevron, on average, increases by 0.09178. In plain terms, a more conservative Supreme Court has been more historically correlated with a higher agency win rates in cases that cite Chevron Deference.

However, the r squared value is 0.1238123. An r-squared value closer to one demonstrates a stronger level of correlation. Thus, because the the r squared value is relatively low (close to zero), it shows there is a relatively weak relationship between ideological slant of the supreme court and average agency win rate for a given term. Because the correlation is fairly weak, it is unlikely this relationship should be interpreted causally. This is because, on average, the model prescribed above is a weak predictor of the dependent variable.


But, it should be noted that the p-value for the relationship above is 0.0353315. At a significance level of 0.05, this satisfies the condition for a statistically significant relationship. With a low r squared and a statistically significant p-value, we can interpret that the above model does not explain the relationship in the data paticularly well, but it is signficant in the sense that it is better than having no model at all.

## Conclusion

